\begin{thebibliography}{9}

\bibitem{ltb-2017}
 Landlord and Tenant Board. (2017, May). Brochure: A Guide to the Residential Tenancies Act. Retrieved September 29, 2017, from \href{http://www.sjto.gov.on.ca/documents/ltb/Brochures/Guide\%20to\%20RTA\%20(English).html}{sjto.gov.on.ca}

\bibitem{^gee-2017}
 Gee, M. (2017, April 20). Rent control isn't the solution to Ontario's housing problem. Retrieved September 12, 2017, from \href{https://beta.theglobeandmail.com/news/toronto/rent-control-isnt-the-solution-to-ontarios-housing-problem/article34753102/}{theglobeandmail.com}

\bibitem{^tal-2017}
 Tal, B. (2017, April 4). Rent Control—The Wrong Medicine. Retrieved September 12, 2017, from \href{https://economics.cibccm.com/economicsweb/cds?ID=2595&TYPE=EC_PDF}{economics.cibccm.com}

\bibitem{selley}
 Selley, C. (2017, March 21). Chris Selley: Rent control is a bad solution to the wrong problem. Retrieved September 12, 2017, from \href{http://nationalpost.com/news/toronto/chris-selley-rent-control-is-a-bad-solution-to-the-wrong-problem}{nationalpost.com}

\bibitem{thesun}
 Lafleur, S., \& Filipowicz, J. (2017, April 15). Rent controls wrong answer to housing crisis. Retrieved September 12, 2017, from \href{http://www.torontosun.com/2017/04/15/rent-controls-wrong-answer-to-housing-crisis}{torontosun.com}

\bibitem{cbc-squeeze}
  McGillivray, K. (2017, April 05). Ontario second-worst economy for young people in Canada: report. Retrieved September 12, 2017, from \href{http://www.cbc.ca/news/canada/toronto/generation-squeeze-ontario-economy-1.4054589}{cbc.ca}
  
\bibitem{cbc-martin}
  Martin, S. (2017, February 22). No fixed address: How I became a 32-year-old couch surfer. Retrieved September 29, 2017, from \href{http://www.cbc.ca/news/canada/toronto/no-fixed-address-how-i-became-a-32-year-old-couch-surfer-1.3985771}{cbc.ca}

\bibitem{tgam-jaafari}
  Jaafari, J. D. (2017, April 14). Toronto tenants crushed by rent hike exemptions. Retrieved September 29, 2017, from \href{https://beta.theglobeandmail.com/news/toronto/toronto-tenants-crushed-by-rent-hike-exemptions/article33806825/}{theglobeanmail.com}

\bibitem{mercer-2017}
  Mercer, G. (2017, May 13). Tenants sound alarm on creeping rents. Retrieved October 31, 2017, from \href{https://www.therecord.com/news-story/7312772-tenants-sound-alarm-on-creeping-rents/}{therecord.com}

\bibitem{toronto-2006}
  City of Toronto (2006) \href{https://www1.toronto.ca/city_of_toronto/social_development_finance__administration/files/pdf/housing_rental.pdf}{Rental Housing Supply and Demand Indicators}. Toronto City Planning and Policy Research

\bibitem{^smith-1983}
 Smith, L. B. (1983). The Crisis in Rental Housing: A Canadian Perspective. The Annals of the American Academy of Political and Social Science, 465(1), p. 3-4, 58-75. doi:10.1177/0002716283465001006

\bibitem{^smith-1988}
  Smith, L. B. (1988). An economic assessment of rent controls: The Ontario experience. The Journal of Real Estate Finance and Economics, 1(3), 217-231. doi:10.1007/bf00658918A

\bibitem{^miron-1995}
 Miron, J. R. (1995). Private Rental Housing: The Canadian Experience. Urban Studies, 32(3), 579-604. doi:10.1080/00420989550012988

\bibitem{rent-control-city-report}
  Staff report for information on Tenant Issues Related to the RTA 1 (Rep. pg 2-3). (2013, October 15). Retrieved August 28, 2017, from \href{http://www.toronto.ca/legdocs/mmis/2013/ex/bgrd/backgroundfile-63467.pdf}{toronto.ca}

\bibitem{^smith-2003}
 Smith, L. B. (2003). Intertenancy Rent Decontrol in Ontario. Canadian Public Policy / Analyse de Politiques, 29(2), 213. doi:10.2307/3552456

\bibitem{^brescia-2005}
 Brescia, V. (2005). The Affordability of Housing in Ontario: Trends, Causes, Solutions (p. 31).

\bibitem{canada-2017}
Government of Canada (2017, May 01). Historical Age Pyramid. Retrieved October 20, 2017, from [statcan.gc.ca](http://www12.statcan.gc.ca/census-recensement/2016/dp-pd/pyramid/pyramid.cfm?geo1=35\&type=1)

\bibitem{^lind-2003}
Lind, H. (2003). Rent regulation and new construction: With a focus on Sweden 1995-2001. Swedish Economic Policy Review, (10), 135-167.

\bibitem{^black-2012}
 Black, J. (September 2012). The Financing and Economics of Affordable Housing: Development Incentives and Disincentives to Private-Sector Participation (p. 8-9, 20), Cities Centre, University of Toronto

\bibitem{^black-2012-2}
 Black (2012, p. 8-9)

\bibitem{^hanes-2008}
  Hanes, T. (2008, February 23). Happy 40th. Retrieved September 18, 2017, from \href{https://www.thestar.com/news/2008/02/23/happy_40th.html}{thestar.com}

\bibitem{^pomeroy-2001}
 Pomeroy, S. (2001, October). Toward a Comprehensive Affordable Housing Strategy for Canada (p. 15).

\bibitem{^lampert-2016}
Lampert, G. (2016). Encouraging Construction and Retention of Purpose-Built Rental Housing in Canada: Analysis of Federal Tax Policy Options (p. 6). 

\bibitem{^hswg-2001}
 Housing Supply Working Group. (2001, May) Affordable Rental Housing Supply: The Dynamics of the Market and Recommendations for Encouraging New Supply, p. 9-14, 17-18

\bibitem{^frpo-2015}
 Federation of Rental-housing Providers of Ontario (2015) Removing Barriers to New Rental Housing in Ontario (p. 18)

\bibitem{^crook-1998} Crook, T. (1998) The Supply of Private Rented Housing in Canada. Netherlands Journal of of Housing and the Built Environment, 13(3), 339

\bibitem{^budget-1971}
 Benson, E. J. (1971, June 18) Budget speech delivered by the Honourable E. J. Benson Minister of Finance and Member of Parliament for Kingston and The Islands. Retrieved from [budget.gc](http://www.budget.gc.ca/pdfarch/1971-sd-eng.pdf)

\bibitem{^clayton-1998}
 Clayton, F. (1998, November) Economic Impact of Federal Tax Legislation on the Rental Housing Market in Canada (p. i, 9-12) Canadian Federation of Apartment Associations - FIX THIS TITLE COS LATEX DOESNT TAKE UNICODE %Fédération Canadienne des Associations de Propriétaires Immobiliers

\bibitem{^enemark-2017}
  Enemark, T. (2017, July 06). Fastest Way to More Rental Housing? Tax Changes. Retrieved September 12, 2017, from \href{https://thetyee.ca/Opinion/2017/07/06/Tax-Changes-More-Rental-Housing/}{thetyee.ca}

\bibitem{^steele-1993}
  Steele, M. (1993) Conversions, Condominiums and Capital Gains: The Transformation of the Ontario Rental Housing Market. Urban Studies 30(1), 103-126

\bibitem{^crook-1998-2} Crook (1998, p. 335-336)

\bibitem{^goldberg-1985}
 Goldberg, M. A., \& Mark, J. H. (1985). The Roles of Government in Housing Policy A Canadian Perspective and Overview. Journal of the American Planning Association, 51(1), 35-36. doi:10.1080/01944368508976798

\bibitem{^suttor-2009}
 Suttor, G. (2009). Rental Paths from Postwar to Present: Canada Compared (p. 39-40, Research Paper 218). Toronto: Cities Centre University of Toronto.

\bibitem{^clayton-1998-2}
 Clayton (1998, p. 9, 32-33)
  

\bibitem{^clayton-1998-3}
 Clayton (1998, p. 14, 28) 

\bibitem{^morales-2017}
  Morales, M. (2017, August 14). Why Seattle Builds Apartments, but Vancouver, BC, Builds Condos. Retrieved September 13, 2017, from \href{http://www.sightline.org/2017/08/14/why-seattle-builds-apartments-but-vancouver-bc-builds-condos/}{sightline.org}

% part two

\bibitem{^arnott-1995}
   Arnott, R. (1995). Time for Revisionism on Rent Control? Journal of Economic Perspectives,9(1), 100-102, 112-115. doi:10.1257/jep.9.1.99

\bibitem{^lind-2001}
  Lind, H. (2001). Rent Regulation: a Conceptual and Comparative Analysis. European Journal of Housing Policy, 1(1), 41-57. doi:10.1080/14616710110036436

\bibitem{^nycrgb-2016}
  New York City Rent Guidelines Board. (2016, September 23). Rent Control FAQ. Retrieved October 30, 2017, from \href{http://www.housingnyc.com/html/resources/faq/rentcontrol.html}{housingnyc.com}

\bibitem{^curbed-2017}
  Nonko, E. (2017, August 28). Rent control vs. rent stabilization in NYC, explained. Retrieved October 30, 2017, from \href{https://ny.curbed.com/2017/8/28/16214506/nyc-apartments-housing-rent-control}{ny.curbed.com}

\bibitem{^arnott-2003}
  Arnott, R (2003). Tenancy Rent Control. Swedish Economic Policy Review, (10), 98

\bibitem{^macfarlane-2001}
  Mcfarlane, A. (2001). Rent Stabilization and the Long-Run Supply of Housing. SSRN Electronic Journal. doi:10.2139/ssrn.283336

\bibitem{^black-2012-4}
  Black, J. (September 2012). The Financing and Economics of Affordable Housing: Development Incentives and Disincentives to Private-Sector Participation (p. 8-12), Cities Centre, University of Toronto

\bibitem{^tait-2017}
  Tait, S. (2017, April 30). Ontario’s Fair Housing Plan \& Purpose-Built Rental Development. Retrieved September 14, 2017, from [linkedin.com](https://www.linkedin.com/pulse/ontarios-fair-housing-plan-purpose-built-rental-development-sean-tait/)

\bibitem{^marr-2017}
  Marr, G. (2017, July 18). Rent controls, no worries. Developer unveils major apartment complex in downtown Toronto. Retrieved September 14, 2017, from [business.financialpost.com](http://business.financialpost.com/real-estate/property-post/one-of-the-best-assets-classes-major-landlord-expects-rental-apartment-demand-to-remain-strong-despite-rent-controls)

\bibitem{^grant-2011}
  Grant, H. (2011, January 31). An Analysis of Manitoba’s Rent Regulation Program and the Impact on the Rental Housing Market. Manitoba Family Services and Consumer Affairs

\bibitem{^ambrosius-2015}
   Ambrosius, J. D., Gilderbloom, J. I., Steele, W. J., Meares, W. L., \& Keating, D. (2015). Forty years of rent control: Reexamining New Jersey’s moderate local policies after the great recession. Cities: The International Journal of Urban Policy and Planning, 49, 121-133. doi:10.1016/j.cities.2015.08.001

\bibitem{^haines-2017}
  Haines, G. (2017, April 28). Making Cents of Ontario’s Rent Regulation: Assessing the Financial Impact of Rent Control and Developer Incentives. Ryerson City Building Institute.

\bibitem{^ambrosius-2015-2}
  Ambrosius et al (2016, p. 129-130)

\bibitem{^grant-2011-2}
  Grant (2011, p. 24)

\bibitem{^denton-1993}
  Denton, F. T., Feaver, C. H., Muller R. A., Robb, A. L., Spencer, B. G. (1993). Testing Hypotheses About Rent Control: Final Report. Ottawa: CMHC.

\bibitem{^lazzarin-1990}
  Lazzarin, C. C. (1990). Rent Control and Rent Decontrol in British Columbia: A Case Study of the Vancouver Rental Market, 1974 to 1989. University of British Columbia.

\bibitem{^sims-2007}
  Sims, D. P. (2007). Out of control: What can we learn from the end of Massachusetts rent control? Journal of Urban Economics, 61(1), 129-151. doi:10.1016/j.jue.2006.06.004

\bibitem{^diamond-2017}
   Diamond, R., McQuade, T., \& Qian, F. (2017). The Effects of Rent Control Expansion on Tenants, Landlords, and Inequality: Evidence from San Francisco.

\bibitem{^mclaughlin-2016}
  McLaughlin, R. (2016, August 31). Rich City, Poor City: How Housing Supply Drives Regional Economic Inequality. Retrieved October 31, 2017, from [trulia.com](https://www.trulia.com/blog/trends/rich-city-poor-city/)

\bibitem{^torres-2017}
   Torres, B. (2017, April 28). Housing’s tale of two cities: Seattle builds, S.F. lags. Retrieved October 30, 2017, from [bizjournals.com](https://www.bizjournals.com/sanfrancisco/news/2017/04/28/san-francisco-seattle-housing-production-pipelines.html)

\bibitem{^hswg-final}
  Lampert, G., Pomeroy, S., \& Bradley, R. (2002). Creating a Positive Climate for Rental Housing Development Through Tax and Mortgage Insurance Reforms: The Second Report of the Housing Supply Working Group (p. 7). Toronto: Ministry of Municipal Affairs and Housing.

\bibitem{^brescia-2005-3}
  Brescia (2005, p. 7-8)

\bibitem{^brescia-2005-2}
   Brescia (2005, p. 27)

% part three
\bibitem{^jenkins-2009}
  Jenkins, B. (2009, January) Rent Control: Do Economists Agree? Econ Journal Watch, 6(1), 73-112

\bibitem{^arnott-2003-2}
  Arnott (2003, p. 111)

\bibitem{^yee-1989}
  Yee, G. (1989) Rationales for Tenant Protection and Security of Tenure, Journal of Law and Social Policy (5), p. 48, 50

\bibitem{^economist-1998}
   The morning after. (1998, May 02). Retrieved October 31, 2017, from [economist.com](http://www.economist.com/node/161526)

\bibitem{^hulchanski-laissez}
  Hulchanski (1984, p. 27)

\bibitem{^brescia-2005-4}
  Brescia (2005, p. 26)

\bibitem{^suttor-abridged}
  Suttor, G. (2016) Canadian Social Housing: Policy Evolution and Program Periods, p. 12

\bibitem{^hulchanski-bulletin38}
  Hulchanski, J. D. (2007, September) Canada’s Dual Housing Policy Assisting Owners, Neglecting Renters, University of Toronto Centre for Urban and Community Studies

\bibitem{^yorke-2015}
  Yorke, B. (2015, December 10). The Tenant Movement in B.C. from 1968 to 1978. Retrieved September 20, 2017, from [themainlander.com](http://themainlander.com/2012/11/09/the-tenant-movement-in-b-c-from-1968-to-1978/)

\bibitem{^hulchanski-tenant-rights}
   Hulchanski (1984, p. 74-75)

 \bibitem{^hulchanski-1984}
    Hulchanski, J. D. (1984) Market Imperfections and the Role of Rent Regulations in the Residential Rental Market, Toronto: Ontario Commission of Inquiry into Residential Tenancies, Research Study No. 6.

\bibitem{^mcardle-2017}
  McArdle, M. (2017, June 16). Beware of Blaming Government for London Tower Fire. Retrieved September 12, 2017, from [bloomberg.com](https://www.bloomberg.com/view/articles/2017-06-16/beware-of-blaming-government-for-london-tower-fire)

\bibitem{^fire-deaths}
  Fire Death Rates. (2016, December). Retrieved September 12, 2017, from [mcscs.jus.gov.on.ca](https://www.mcscs.jus.gov.on.ca/english/FireMarshal/MediaRelationsandResources/FireStatistics/OntarioFatalities/FireDeathRate/stats\_death\_rate.html)

\bibitem{^cowen-2007}
  Cowen, T. (2007, December). Public Goods. Retrieved November 02, 2017, from [econlib.org](http://www.econlib.org/library/Enc/PublicGoods.html)

\bibitem{^galea-2016}
  Quote is by John Kenneth Galbraith via Galea, S. (2016, January 10). Public Health as a Public Good | SPH | Boston University. Retrieved November 02, 2017, from [bu.edu](https://www.bu.edu/sph/2016/01/10/public-health-as-a-public-good/)

\bibitem{^desmond}
   Desmond, M. (2016). Evicted: Poverty and Profit in the American City. Crown Books. p. 70, 296-298.

\bibitem{^mcgrath-2017}
  McGrath, J. (2017, July 14). There's life outside Toronto, but leaving the city is harder than it used to be. Retrieved September 13, 2017, from [tvo.org](http://tvo.org/article/current-affairs/the-next-ontario/theres-life-outside-toronto-but-leaving-the-city-is-harder-than-it-used-to-be)

\bibitem{^waldegrave-2016}
   Waldegrave, C., FIX LATEX UNICODE %Urbanová, M. (2016, November) Social and Economic Impacts of Housing Tenure. New Zealand Housing Foundation.

\bibitem{^tal-rent-2017}
  Tal, B. (2017, March 15). GTA Housing—Rent Must be Part of the Solution. Retrieved September 1, 2017, from [economics.cibccm.com](https://economics.cibccm.com/economicsweb/cds?ID=2475\&TYPE=EC\_PDF)

\bibitem{^census-2011}
  Statistics Canada (2017, May 3) Census concepts: A profile of the City of Toronto, Toronto Census 2016 Community Planning Session.

\bibitem{^carrick-2017}
  Carrick, R. (2017, November 02). Accept it, your children may grow up to be renters. Retrieved November 02, 2017, from [theglobeandmail.com](https://beta.theglobeandmail.com/globe-investor/personal-finance/household-finances/accept-it-your-children-may-grow-up-to-be-renters/article36806999/)

\bibitem{^suttor-2017}
  Suttor, G., \& Leon, S. (2017, October 27). 6 Toronto Rental Housing Highlights in the 2016 Census. Retrieved November 7, 2017, from [wellesleyinstitute.com](http://www.wellesleyinstitute.com/housing/6-toronto-rental-housing-highlights-in-the-2016-census)


\end{thebibliography}

